---
title: "ERC Analysis Walkthrough"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: yes
  html_document:
    css: custom.css
    toc: yes
authors: Jordan Little, Guillermo Hoffmann Meyer
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) 
```
\newpage
## Installing and loading ERC

Make sure to visit the [**installation page**](https://github.com/nclark-lab/erc/blob/main/install.md) to get the code and prerequisite packages before following this tutorial.

# ERC Pipeline Walkthrough
You can follow along in R or RStudio, or you can read along in **runERC.R**. First, if you are using RStudio, you need to set your directory. Change the string in `setwd` to the directory your runERC.R file is in, generally the repository you cloned/downloaded.

```{r}
setwd("~/Documents/GitHub/erc")
```

Once you set your directory, you can source the relevant packages and ERC files.

```{r results='hide', message = FALSE, warning = FALSE, cache = FALSE}
require(devtools)
remotes::install_github("ms609/TreeTools")
source("ERC_functions.R")
source("ERC.R") 
Rcpp::sourceCpp("cppFuncs.cpp")
```

## File setup

To run the workflow, you need to delineate two things: the tree file to read in, and your output file. Find the path for the tree file, and choose a name (and optionally a path) for your output file. Here we set `treefile` and `outputfile` accordingly.

```{r, cache = TRUE}
treefile = "physical_interaction_paper/domains_trees.tre"
outputfile = "out.RDS"

```

## Workflow

Now that you've selected your file names, you can begin running the main functions. For a detailed description of what they do and their parameters, visit the [functions page](https://github.com/nclark-lab/erc/blob/main/functions.md). First, your tree file is read in using `readTrees`. The trees are then transformed (via a square root transform) with `transformPaths`. 

```{r message = FALSE, warning = FALSE, cache = TRUE}
trees=readTrees(treefile)

compTrees = transformPaths(trees, transform = "sqrt",impute = F)
```

Above is a plot of the tree paths before and after a square root transform.

Next, Relative Evolutionary Rates (RERs) are calculated by `coreGetResiduals`, and finally those rates are formatted into a matrix with `getRMat`. `getAllResiduals` (commented out below) has the same output as running these three together, but here we do not use it because we will later need the compTrees object.

```{r message=FALSE, warning=FALSE, cache=TRUE}
compResid = coreGetResiduals(compTrees, n.pcs=0)
residuals = getRMat(compResid, all = T, rmatweights = compTrees$weights)

# Wrapper function: (you still need to separately get compTrees
#                    for the later clusterList function)
# residuals = getAllResiduals(trees, impute=F, n.pcs=0, all=T)
```



Finally, we get a list of gene clusters for the ERC function.
```{r message = FALSE, warning = FALSE, cache = TRUE}
clusterList = getClusterList(compTrees)
```

## ERC function

Finally, you can compute the ERC values for your trees. You can tune many parameters (also visible on the [**functions page**](https://github.com/nclark-lab/erc/blob/main/functions.md)), but the main ones you want to worry about are below:
- Here is where you will edit the threshold you want, `minSp` is the number of species two genes have to share
- If you only want to run a few genes you can set the parameter `doOnly = c("genea","geneb")`
- If you want the plot of the RERs set `plot = T` (I would only recommend doing this for a few genes because it uses up a lot of space)
```{r}
corres=computeERC(residuals, compTrees, clusterListOutput = clusterList,
                  minSp = 15, saveFile = outputfile)
```

## Fisher transformation

After you create your ERC matrices, we recommend Fisher transforming them. This creates a single matrix taking into account the two matrices of the `corres` object: the correlation matrix and the matrix of observation/branch counts for each correlation. We also make it symmetrical here, but if your matrix is too large you may just want to make subsets symmetrical as you need them.

```{r}
ft_data = fisherTransform(corres)

#makes the matrix symmetrical
sym_ft = make_symmetric(ft_data)
```

Congratulations for making it to the end! Now with `ft_data` you have the data we usually operate on. Below we have some sample analysis you can do with it.


# Next Steps:

## Example: examine 10 genes' relations to each other

In this example, we show how to visualize ERC data. We take a sample ten genes, and create a symmetrical ERC matrix of their values (we round the values at the end to make display clearer).

```{r}
genes = c("NSE5_1", "NSE6_3",  "CSE1_3",  "CSE1_1",  "EXO70_1",
          "MCM2_4",  "MDY2_1",  "ATP1_2",  "MCM5_1",  "SEC8_2")
# You could also generate a random sample:
# genes = colnames(ft_data)[sample(1:length(ft_data), 10, replace=FALSE)]


# makes a matrix of the 10 genes against themselves 
# (it can be against different genes too)
ft_filtered = betweencomplex(genes,genes,sym_ft)

ft_filtered = round(ft_filtered,3)
#output
ft_filtered


```

## Another example: non-Fisher transformed data

We can also use our raw ERC correlation data from before the Fisher transformation. Again, we round to simplify the display.

```{r}


filtered = betweencomplex(genes,genes,corres[["cor"]])

sym = round(make_symmetric(filtered),3)

#output
sym
```